---
title: "PCA"
author: "ALSF CCDL - Candace Savonen"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

*Purpose*: This notebook illustrates how to perform Principal Components 
Analysis and plot the scores using ggplot2. 

## 1) Set up libraries

```{r}
# Magrittr pipe
`%>%` <- dplyr::`%>%`

# Load ggplot2 for graphing
library(ggplot2)
```

## 2) Import and set up data
Data downloaded from refine.bio include a metadata tab separated values ("tsv")
file and a data tsv file. 
These Liver gene expression files were obtained from [refine.bio](https://www.refine.bio/) 
by entering and then aggregating by `species`.

```{r Import refine.bio data}
# Put name of data file
data.filename <- file.path("data", "HOMO_SAPIENS.tsv")

# Unzip the file and direct it to the data folder
unzip(paste0(data.filename, ".zip"), exdir = "data")

# Read in data tsv file
df <- readr::read_tsv(data.filename, progress = FALSE) %>% 
  tibble::column_to_rownames('X1') 
# Make the gene column the rownames so the gene names are out of the way for calculations
```

Prep metadata

```{r Prep metadata}
# Metadata file name
metadata.filename <- file.path("data", "metadata_HOMO_SAPIENS.tsv")

# Read in data tsv file
metadata <- readr::read_tsv(metadata.filename)
```

Check sample order

```{r}
# Check if the samples are in the same order
all.equal(metadata$refinebio_accession_code, colnames(df))

# Put samples in the same order in the dataframe
df <- df[, metadata$refinebio_accession_code]
```

## 3) Perform Principal Components Analysis.

```{r}
# Calculate PCA
pca <- prcomp(t(df))

# Make into dataframe for plotting with ggplot2
pca.df <- data.frame(pca$x[, 1:2], 
                     dataset = metadata$series_id)
```

Plot these PC scores.
We will likely see a difference between the datasets.

```{r}
ggplot(pca.df, aes(x = PC1, y = PC2, color = metadata$refinebio_platform,
       shape = metadata$series_id)) +
  geom_point() + 
  theme(legend.position = "none") +
  ggtitle("PC scores")
```

## 4) Take a look at the gene loadings.

```{r}

```

## 5) Save PC scores to tsv file.

```{r}
pca.df %>% as.data.frame() %>% 
  tibble::rownames_to_column("gene") %>%
  readr::write_tsv(file.path("results", "Liver_PCA_scores.tsv"))
```

Session info: 

```{r Print session info}
# Print session info 
sessionInfo()
```
