---
title: "KEGG pathways: mapping to mouse orthologs with `hcop`"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: J. Taroni for ALSF CCDL
date: 2019
---

## Background

In this module, we use QuSAGE ([Yaari et al. _NAR_. 2013.](https://doi.org/10.1093/nar/gkt660))
for pathway analysis (implemented in the [`qusage` bioconductor package](https://bioconductor.org/packages/release/bioc/html/qusage.html)).

`qusage` allows you to read in gene sets that are in the [GMT format](http://software.broadinstitute.org/cancer/software/gsea/wiki/index.php/Data_formats#GMT:_Gene_Matrix_Transposed_file_format_.28.2A.gmt.29).

[MSigDB](http://software.broadinstitute.org/gsea/msigdb) offers genesets in this format.
[Curated gene sets](http://software.broadinstitute.org/gsea/msigdb/collections.jsp#C2) 
such as [KEGG](https://www.genome.jp/kegg/) are a good starting point for any pathway analysis.

However, MSigDB only distributes human pathways.
If we want to use KEGG Pathways with another species without going through
[KEGG Orthology](https://www.genome.jp/kegg/ko.html), we need to map to
orthologs ourselves.

We'll use the [`hcop` package](https://github.com/stephenturner/hcop) to do 
this.
If you're looking for a little bit more background information (like if you
run into trouble installing `hcop`), check out the notebook in our
[`ortholog-mapping`](https://github.com/AlexsLemonade/refinebio-examples/tree/master/ortholog-mapping) module.

## Setup

Package installation

```{r}
# need read.gmt functionality from qusage
if (!("qusage" %in% installed.packages())) {
  BiocManager::install("qusage", update = FALSE)
}

# we need devtools in order to install the hcop package we will use to
# do the ortholog mapping
if (!("devtools" %in% installed.packages())) {
  install.packages("devtools")
}

# this installs a specific version of hcop
# we pass the commit hash to the ref argument
devtools::install_github("stephenturner/hcop", 
                         ref = "0985fddc91a6ef2308f4800958dfd11c25fe6a98")
```

```{r}
`%>%` <- dplyr::`%>%`
```

```{r}
library(hcop)
```

## KEGG human pathways

We need to download the the MSigDB v6.2 KEGG gene sets that use Entrez gene IDs
and place them at the following path if we have not done so already:

```
gene-sets/c2.cp.kegg.v6.2.entrez.gmt
```

```{r}
# the kegg gmt file should be located in the spot we mention above
kegg_file <- file.path("gene-sets", "c2.cp.kegg.v6.2.entrez.gmt")
# since we do not track this file in our repository, let's check to make sure
# it exists where we expect it and download it if we don't find it
if (!file.exists(kegg_file)) {
  message(paste("KEGG GMT file is not found at", kegg_file, 
                ", downloading now..."))
  # need gene-sets directory
  if (!dir.exists("gene-sets")) {
    dir.create("gene-sets")
  }
  download.file("https://data.broadinstitute.org/gsea-msigdb/msigdb/release/6.2/c2.cp.kegg.v6.2.entrez.gmt", 
                destfile = kegg_file)
}
```

Read in the pathway file

```{r}
kegg_human_list <- qusage::read.gmt(kegg_file)
```

## Conversion from human Entrez ID to mouse symbol

### Human to mouse mapping with `hcop`

`hcop` is designed to work well with `dplyr`, so we'll get all the possible
human Entrez IDs into a `data.frame`.

```{r}
entrez_in_pathways_df <- data.frame(
  human_entrez = as.integer(unique(unlist(kegg_human_list)))
)
```

Join to mouse orthologs by the human Entrez IDs.

```{r}
# Join to mouse orthologs
mouse_ortholog_df <- entrez_in_pathways_df %>% 
  dplyr::inner_join(mouse, by = "human_entrez")
```

### 1:many mapping

For 1:many mappings, we'll pick the one with the most support.
Here, we'll only consider the _number_ of sources without any regard for _what_
those resources are.
Essentially, we weigh all sources equally though they likely have more or less
permissive criteria.

```{r}
# add a column that counts resources
mouse_ortholog_df <- mouse_ortholog_df %>%
  dplyr::rowwise() %>%
  dplyr::mutate(num_resources_support = 
                  length(stringr::str_split(support, 
                                            pattern = ",", 
                                            simplify = TRUE)))

```

To demonstrate how this works, we'll pick a gene that had 1:many mappings
and follow it along.

```{r}
mouse_ortholog_df %>% 
  dplyr::filter(human_entrez == 5631) %>%
  dplyr::select(human_entrez, mouse_entrez, human_symbol, mouse_symbol,
                num_resources_support)
```

We can see that the human gene _PRPS1_/`5631` maps to 4 mouse genes, one of 
which has 10 resources supporting that mapping.

```{r}
# for each unique human entrez id, pick the mapping with the highest number of
# resources supporting it
most_support_df <- mouse_ortholog_df %>%
  dplyr::group_by(human_entrez) %>%
  dplyr::top_n(1, num_resources_support)
```

What happened to `5631`?

```{r}
most_support_df %>% 
  dplyr::filter(human_entrez == 5631) %>%
  dplyr::select(human_entrez, mouse_entrez, human_symbol, mouse_symbol,
                num_resources_support)
```

We successfully selected the mapping with the highest number of resources
supporting it.

### Conversion of KEGG pathways

For each KEGG pathway we have (currently populated with human Entrez IDs),
we need a new gene set that is comprised of mouse gene symbols.

```{r}
kegg_mouse_list <- 
  lapply(kegg_human_list, 
         function(x) {
           dplyr::filter(most_support_df, human_entrez %in% x) %>%
             dplyr::pull(mouse_symbol)
         })
```

Do the results this seem reasonable?
Let's pick the polymerase pathway, where our success should be pretty obvious
from the mouse gene symbols.

```{r}
kegg_mouse_list[[grep("POLYMERASE", names(kegg_mouse_list))]]
```

### Write mouse pathway list in `GMT` format

We need to write this new mouse pathway list to file in [GMT format](http://software.broadinstitute.org/cancer/software/gsea/wiki/index.php/Data_formats#GMT:_Gene_Matrix_Transposed_file_format_.28.2A.gmt.29), as this will facilitate use with downstream pathway analysis.

Briefly, the GMT format has one pathway per line and it follows this pattern:

```
<PATHWAY_NAME>\t<PATHWAY_DESCRIPTION>\t<gene_1_in_pathway>...<gene_n_in_pathway>
```

We've lost the description information because it's removed by 
`qusage::read.gmt`.
The description in `r kegg_file` follows this pattern:

```
http://www.broadinstitute.org/gsea/msigdb/cards/<PATHWAY_NAME>
```

We can pretty easily stick this back in when we write to file with `write()`.

```{r}
# filename we're going to write to
mouse_file <- file.path("gene-sets", 
                        "c2.cp.kegg.v6.2.entrez_mouse_symbol_hcop.gmt")

# if you run this again after making changes above, you'd just end up appending
# this to the end of your old file, so we should take steps to get rid of
# the existing file
if (file.exists(mouse_file)) {
  message(paste("Removing old", mouse_file))
  file.remove(mouse_file)
}

# for each pathway, write it to line following GMT format
for (pathway_iter in seq_along(kegg_mouse_list)) {
  # extract the current pathway name
  pathway_name <- names(kegg_mouse_list)[pathway_iter]
  text_to_write <- 
    paste(
      # the name of the pathway
      pathway_name,
      # the description -- this is stripped out by qusage::read.gmt
      paste0("http://www.broadinstitute.org/gsea/msigdb/cards/",
            pathway_name),
      # the gene symbols
      paste(kegg_mouse_list[[pathway_iter]], collapse = "\t"),
      sep = "\t")
  write(text_to_write, mouse_file, append = TRUE)
}
```

We can double check how this went by reading it back in with `qusage::read.gmt`.

```{r}
mouse_read_list <- qusage::read.gmt(mouse_file)
all.equal(mouse_read_list, kegg_mouse_list)
```
