---
title: "Ortholog mapping with `hcop`"
author: ALSF CCDL 2018
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This notebook demonstrates how you can use the 
[`hcop`](https://github.com/stephenturner/hcop) package to perform ortholog 
mapping for data obtained from refine.bio.

HCOP stands for HGNC Comparison of Orthology Predictions.
You can read more about the package 
[here](https://stephenturner.github.io/hcop).

## 1) Install `hcop`

We'll need to use the 
[`devtools`](https://cran.r-project.org/web/packages/devtools/index.html) 
package to install `hcop` from Github.

`devtools` can be installed using the instructions below.
Note that this will first check if `devtools` is installed and install it
if it is not.

```{r Install devtools}
if (!("devtools" %in% installed.packages())) {
  install.packages("devtools")
}
```

_Note:_ `devtools` requires `git2r` which we've encountered trouble installing
before.
Specifically, `git2r` requires the [`libgit2`](https://libgit2.org/) and 
[`zlib`](https://zlib.net/) system libraries.
If you're using a Docker image from the 
[Rocker project](https://www.rocker-project.org/) without these dependencies 
(e.g., `rocker/rstudio`), follow 
[these instructions](https://github.com/rocker-org/rocker/wiki/Using-the-RStudio-image#dependencies-external-to-the-r-system).

Now we'll install `hcop` from Github.
You can control what version of the package is installed by using the `ref` 
command of `devtools::install_github`.
Here, we'll use the most recent commit at the time that we are putting together
this example.

```{r Install hcop}
devtools::install_github("stephenturner/hcop", 
                         ref = "0985fddc91a6ef2308f4800958dfd11c25fe6a98")
```

## 2) Mapping zebrafish ensembl gene ids to human symbols

`hcop` supports mapping between zebrafish and human identifiers. 
Here, we'll use zebrafish data from refine.bio and annotate it with human gene
symbols.

In keeping with the `hcop` documentation, we'll use tidyverse packages 
(e.g., [`dplyr`](https://dplyr.tidyverse.org/) and 
[`readr`](https://readr.tidyverse.org)) below.

```{r Load hcop and dplyr libraries}
library(hcop)
library(dplyr)
```

Let's read in the tsv file from refine.bio. 
We'll convert _all_ identifiers in the file, rather than selecting a few.

```{r Read in zebrafish gene expression data}
exprs.df <- readr::read_tsv("GSE12955.tsv")
```

refine.bio data uses Ensembl gene identifiers, which will be in the first 
column.

```{r Examine identifiers head}
head(exprs.df[[1]])
```

We can see that this also includes Affymetrix control probes in this case.
These will get dropped using `inner_join` below, because they will not
map to a human gene symbol.

In zebrafish, Ensembl gene identifiers begin with `ENSDARG`.

```{r Examine ENSDARG}
head(exprs.df[[1]][grepl("ENSDARG", exprs.df[[1]])])
```

Let's do the conversion itself!
We'll use the `zebrafish` data from `hcop`; this is what `zebrafish` looks like.

```{r zebrafish}
zebrafish
```

We're interested in the `human_symbol` and `zebrafish_ensembl` columns
specifically.

```{r Conversion}
human.symbol.df <- zebrafish %>%
  # we'll want to subset zebrafish to only the columns we're interested in
  select(zebrafish_ensembl, human_symbol) %>%
  # only retain the genes from the expression matrix that are in zebrafish data
  inner_join(exprs.df, by = c("zebrafish_ensembl" = "X1"))
```

Here's what the new data.frame looks like:

```{r Head of converted df}
head(human.symbol.df, 25)
```

```{r Write to file}
readr::write_tsv(human.symbol.df, "GSE12955_mapped_to_human_gene_symbol.tsv")
```

There are some things to note:

* Multiple zebrafish Ensembl gene ids map to the same human symbol which results
in duplicated human gene symbols. Some downstream tools may need this to be
resolved.
* If a zebrafish Ensembl gene id maps to multiple human symbols, the gene 
expression values get duplicated. Let's look at the `ENSDARG00000069142` 
example below.

```{r}
human.symbol.df %>%
  filter(zebrafish_ensembl == "ENSDARG00000069142")
```

