---
title: "Clustering Heatmap - RNA-Seq"
author: "CCDL for ALSF"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

*Purpose*: This notebook illustrates one way that you can use harmonized data 
from refine.bio in downstream analyses.

TODO: This introduction section will be revamped when the intro to an example
analysis (#122: https://github.com/AlexsLemonade/refinebio-examples/pull/122)
gets more polished/determined.

## Install libraries
This script uses the R package `DESeq2` for normalization and the R package
`pheatmap` for clustering and creating a heatmap.
<br/> <i> Citation:</i> Love, M.I., Huber, W., Anders, S. Moderated estimation of fold change and dispersion for RNA-Seq data with DESeq2 Genome Biology 15(12):550 (2014)
<br/> <i> Citation:</i> Raivo Kolde (2019). pheatmap: Pretty Heatmaps. R package version 1.0.12. https://CRAN.R-project.org/package=pheatmap

```{r Install and attach the pheatmap and DESeq2 libraries}
if (!("pheatmap" %in% installed.packages())) {
  # Install pheatmap
  install.packages("pheatmap", update = FALSE)
}

if (!("DESeq2" %in% installed.packages())) {
  # Install DESeq2
  BiocManager::install("DESeq2", update = FALSE)
}
```

Attach the `pheatmap` and `DESeq2` libraries:

```{r}
# Attach the `pheatmap` library
library(pheatmap)

# Attach the `DESeq2` library
library(DESeq2)

# Magrittr pipe
`%>%` <- dplyr::`%>%`

# Set the seed so our results are reproducible:
set.seed(12345)
```

Create output folder.

```{r}
# Define the file path to the plots directory
plots_dir <- "plots" # Replace with path to desired output plots directory

# Create the plots folder if it doesn't exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

## Import and set up data
Data downloaded from refine.bio include a metadata [tab separated values ("tsv")]
(https://en.wikipedia.org/wiki/Tab-separated_values) file and a data tsv file.
This chunk of code will read the both tsv files and add them as data.frame
objects to your environment.

The data that we downloaded from refine.bio for this analysis has 19 samples
(obtained from 19 mice with acute myeloid leukemia (AML)), containing
RNA-sequencing results for types of AML under controlled treatment conditions.

TODO: Add link to the `getting started` section, specifically to point out how
one can download the dataset.

Let's import the metadata and count data files.

```{r Import data from .tsv files}
# Read in metadata tsv file
metadata <- readr::read_tsv(file.path(
  "data", # Replace with path to your metadata file
  "metadata_SRP070849.tsv" # Replace with the name of your metadata file
))
  # # We only need to select the three variables from the metadata that we will
  # # need for this notebook
  # dplyr::select(refinebio_accession_code, refinebio_title, refinebio_treatment)

# Read in data tsv file
df <- readr::read_tsv(file.path(
  "data", # Replace with path to your data file
  "SRP070849.tsv" # Replace with the name of your data file
)) %>%
  # Here we are going to store the gene IDs as rownames so that we can have a
  # numeric matrix to perform calculations on later 
  tibble::column_to_rownames("Gene")
```

Let's take a look at the metadata object that we read into the R environment.

```{r}
head(metadata)
```

Now let's ensure that the metadata and data are in the same sample order. 

```{r}
# Make the data in the order of the metadata
df <- df %>% dplyr::select(metadata$refinebio_accession_code)

# Check if this is in the same order
all.equal(colnames(df), metadata$refinebio_accession_code)
```

## How are our samples clustering together?

### Create a DESeqDataset
We will be using the `DESeq2` package for normalization, which requires us to
format our data into a `DESeqDataSet` object. We turn the data.frame (or matrix)
into a [`DESeqDataSet` object](TODO: Add link to header section) and specify
which variable labels our experimental groups using the `design` argument. 

```{r Create a DESeqDataset}
# The `DESeqDataSetFromMatrix` function needs the values to be converted to
# integers
df <- df %>%
  # Mutate numeric variables to be integers
  dplyr::mutate_if(is.numeric, round)

# Create a `DESeqDataSet` object
dds <- DESeqDataSetFromMatrix(
  countData = df, # This is the data.frame with the counts values for all replicates in our dataset
  colData = metadata, # This is the data.frame with the annotation data for the replicates in the counts data.frame
  design = ~ refinebio_treatment # Replace with an appropriate design variable for your analysis
)
```

### Define a minimum counts cutoff
We want to filter out the genes that have not been expressed or that have
low expression counts because we want to remove any possible noise from our data
before we normalize the data and create our heatmap. We are going to do some
pre-filtering to keep only genes with 10 or more reads total. Note that rows
represent gene data and the columns represent sample data in our dataset.

```{r Plot distribution of counts}
# Let's take a look at the distribution of counts first

# Sum up the counts for each gene -- which means we want the sum of counts for
# each row as the rows represent gene data and the columns represent sample data
# in our dataset
count_sums <- data.frame(count = rowSums(counts(dds)))

# Create a density plot using the sums
ggplot2::ggplot(count_sums, ggplot2::aes(x = count)) +
  ggplot2::geom_density() +
  ggplot2::geom_vline(ggplot2::aes(xintercept = 10), # This line represents the current cutoff of 10
                      color = "blue",
                      linetype = "dashed") +
  ggplot2::geom_vline(ggplot2::aes(xintercept = mean(count)), # This line represents the mean count value
                                   color = "red",
                                   linetype = "dashed")
```

```{r Define a minimum counts cutoff}
# Define a minimum counts cutoff and filter `DESeqDataSet` object to include
# only rows that have counts above the cutoff
genes_to_keep <- rowSums(counts(dds)) >= 10
dds <- dds[genes_to_keep, ]
```

### Perform DESeq2 normalization
We are going to use the `rlog` function from the `DESeq2` package to normalize
the data. 
TODO: Add a note/link to section in documentation on why we use `rlog` vs `vst`
and `normTransform`.

```{r Perform DESeq2 normalization}
# Normalize the data in the `DESeqDataSet` object using the
# `rlog` function from the `DESEq2` R package
dds_norm <- rlog(dds)
```

### Choose genes of interest
Although you may want to create a heatmap including all of the genes in the set,
alternatively, the heatmap could be created using only genes of interest. 
For this example, we will sort genes by variance, but there are many alternative
criterion by which you may want to sort your genes <i>eg</i> fold change,
t-statistic, membership to a particular gene ontology, so on. 

```{r Choose genes of interest}
# Calculate the variance for each gene
variances <- apply(assay(dds_norm), 1, var)

# Determine the upper quartile variance cutoff value
upper_var <- quantile(variances, 0.75)

# Subset the data choosing only genes whose variances are in the upper quartile
df_by_var <- data.frame(assay(dds_norm)) %>%
  dplyr::filter(variances > upper_var)
```

### Create a heatmap
To further customize the heatmap, see a vignette for a guide at this [link](https://slowkow.com/notes/pheatmap-tutorial/).

```{r Create a heatmap}
# Create and store the technical replicates heatmap object
pheatmap <-
  pheatmap(
    df_by_var,
    cluster_rows = FALSE, # Replace with `TRUE` to cluster heatmap by rows (genes in this case)
    cluster_cols = FALSE, # Replace with `TRUE` to cluster heatmap by columns (samples in this case)
    main = "Heatmap",
    colorRampPalette(c("deepskyblue",
                       "black",
                       "yellow"))(25
                       ),
    width = 15
  )
```

Now let's add some annotation bars to our heatmap.

### Add annotation bars
We are going to manipulate the metadata and add variables with the information
that we would like to use to annotate the heatmap.

```{r Prepare the metadata for annotation}
# Let's prepare the annotation data.frame for the uncollapsed `DESeqData` set
# object which will be used to create the technical replicates heatmap
annotation_df <- metadata %>%
  # Create a variable to store the cancer type information
  dplyr::mutate(
    cancer_type = dplyr::case_when(
      startsWith(refinebio_title, "TET2") ~ "TET2",
      startsWith(refinebio_title, "IDH2") ~ "IDH2",
      startsWith(refinebio_title, "WT") ~ "WT",
      TRUE ~ "unknown")
  ) %>%
  # We want to select the variables that we want for annotating the technical
  # replicates heatmap
  dplyr::select(refinebio_accession_code,
                cancer_type,
                refinebio_treatment) %>%
  # The `pheatmap` function requires that the rownames of our annotation object
  # matches the column names of our `DESeaDataSet` object
  tibble::column_to_rownames("refinebio_accession_code")
```

#### Create annotated heatmap
You can create an annotated heatmap by providing our annotation object to the
`annotation_col` argument of the `pheatmap` function.

```{r Create annotated heatmap}
# Create and store the annotated heatmap object
pheatmap_annotated <-
  pheatmap(
    df_by_var,
    cluster_rows = FALSE, # Replace with `TRUE` to cluster heatmap by rows (genes in this case)
    cluster_cols = FALSE, # Replace with `TRUE` to cluster heatmap by columns (samples in this case)
    annotation_col = annotation_df,
    main = "Annotated Heatmap",
    colorRampPalette(c("deepskyblue",
                       "black",
                       "yellow"))(25
                       ),
    width = 15
  )
```

### Save heatmap as a png
You can easily switch this to save to a jpeg or tiff by changing the function 
and file name within the function to the respective file suffix.

```{r Save annotated heatmap as a png}
# Open a png file
png(file.path(
  plots_dir,
  "HeatmapSRP070849_annotated.png" # Replace file name with a relevant output plot name
))

# Print your heatmap
pheatmap_annotated

# Close the png file:
dev.off()
```

### Create clustered heatmap and save
You can create a clustered heatmap by telling the `pheatmap` function to cluster
the rows and/or columns.

```{r Create clustered heatmap}
# Create and store a heatmap object with clustered samples and genes
pheatmap_clustered <-
  pheatmap(
    df_by_var,
    cluster_rows = TRUE, # Replace with `TRUE` to cluster heatmap by rows (genes in this case)
    cluster_cols = TRUE, # Replace with `TRUE` to cluster heatmap by columns (samples in this case)
    annotation_col = annotation_df,
    main = "Clustered Heatmap",
    colorRampPalette(c("deepskyblue",
                       "black",
                       "yellow"))(25
                       ),
    width = 15
  )
```

Now let's save our clustered heatmap.

```{r Save clustered heatmap as a png}
# Open a png file
png(file.path(
  plots_dir,
  "HeatmapSRP070849_clustered.png" # Replace file name with a relevant output plot name
))

# Print your heatmap
pheatmap_clustered

# Close the png file:
dev.off()
```

Print session info:

```{r Print session info}
# Print session info
sessionInfo()
```
