---
title: "Comparing Differential Expression Analysis"
author: "ALSF CCDL - Candace Savonen"
output:   
  html_notebook:
    toc: true
    toc_float: true
---

*Purpose*: This notebook takes data processed by refine.bio and compares the
differential expression results to a list of differential expressed genes.

You may obtain differential expression results from your own data or a
collaborator's data.
It is good practice to validate your findings in another dataset that measures a
similar tissue with a similar experimental design.
You can use refine.bio datasets and the steps in this example to quickly check
for agreement with your results.
Ideally when comparing datasets' differential expression results, both results
have been obtained using the same exact pipeline.

Here we will compare differential expression results we've previously obtained
to results from a refine.bio dataset with similar data.

To represent differential expression results you may have obtained, we are using
[Northcott et al, 2012. Nature.](https://www.ncbi.nlm.nih.gov/pubmed/22832581)
using [author-processed data from Gene Expression Omnibus](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE37382).
In another notebook in this module, [`author_processed_DE`](./author_processed_DE.nb.html),
we used `limma` to compare medulloblastoma gene expression data from `SHH`
subgroups versus the other subgroups.
From this we obtained a table of output results (see `data/GSE37382_limma_results.tsv`)
that includes probe IDs, and their associated summary statistics from our
differential expression testing.

In this notebook, we are looking for whether the results we found in the Northcott
data are generalizable to [refine.bio dataset that also measures medulloblastoma subgroups](https://www.refine.bio/experiments/GSE37418/novel-mutations-target-distinct-subgroups-of-medulloblastoma).
These data are originally from [Robinson et al, 2012. Nature.](https://www.ncbi.nlm.nih.gov/pubmed/22722829)
but were downloaded from refine.bio.
This dataset was chosen because, like Northcott data, it also has subgroup
medulloblastoma information that we will use in this notebook for differential
expression testing.

The output from this notebook are two Venn diagrams showing the gene list overlap
between these two datasets (one for up-regulated gene list, and the other for
down-regulated).
In both these datasets, we consider genes that are significantly higher in `SHH`
subgroup samples than in `NonSHH` samples to be `up-regulated`.

There are a number of ways to compare differential expression results, but this
notebook's example takes the simplest approach of looking at overlap of gene lists.
This is partially so that this notebook's approach can be applied to any
differential expression output, provided that the output meets certain criteria
listed below.

### What you will need to apply this dataset to your own results:
At the bare minimum, you will need two gene lists, one for up-regulated and
another for down-regulated genes.

Ideally you will have table of differential expression results that will have:  

1) A gene or probe identifier
2) The direction of the expression differences (often indicated by fold-change)
3) p-values that are corrected for multiple testing (e.g., FDR)  

This example shows one method of summarizing and converting Affymetrix probe IDs
to Ensembl gene IDs, but depending on what type of gene identifiers and results
you have, you may need to do a few conversion steps.

## 1) Set up

```{r}
if (!("limma" %in% installed.packages())) {
  # Install limma
  BiocManager::install("limma", update = FALSE)
}
if (!("VennDiagram" %in% installed.packages())) {
  # Install VennDiagram
  BiocManager::install("VennDiagram", update = FALSE)
}
```

Attach the `limma` library:

```{r message=FALSE}
# We will need this so we can use the pipe: %>%
library(magrittr)

# Attach library
library(limma)
```

Set up output directories if they haven't been set up yet.

```{r}
# Define the file path to the results directory
results_dir <- "results" # Replace with path to desired output results directory

# Make a results directory if it isn't created yet
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

# Define the file path to the plots directory
plots_dir <- "plots" # Replace with path to desired output plots directory

# Make a plots directory if it isn't created yet
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}

# Define the file path to the data directory
data_dir <- "data" # Replace with path to data directory
```

## 2) Import differential expression results from Northcott dataset

Import the `limma` results obtained for the author-processed data for
Northcott et al, 2012.
If you'd like to see how we generated these results with `limma`, see
[`author_processed_DE`](./author_processed_DE.nb.html) in this folder.
Note that these differential expression results come with Affymetrix probe IDs
(our source of gene identifier), `logFC`(our source of direction), and `adj.P.Val`
(our source of significance level).

```{r}
# Load in the summary statistics from limma run
author.full.results <- readr::read_tsv(file.path(
  results_dir,
  "GSE37382_limma_results.tsv" # Replace with name of your own differential expression results file you've generated that you'd like to compare to refine.bio data
))
```

## 3) Convert the differentially expressed gene identifiers to Ensembl gene IDs

In refine.bio all IDs are converted to Ensembl gene IDs in the harmonization
pipeline.
Here we need to convert our Affymetrix transcript cluster IDs into Ensembl gene
IDs so we can compare our datasets.
If you are using this notebook for your own data, and need more information on
how to perform ID conversion, you can look at [`ensembl_id_convert`](./ensembl_id_convert.html)
in this repository for more information.
You can also use the [ID conversion repository prepared by the CCDL.](https://github.com/AlexsLemonade/identifier-refinery)
The pre-made ID conversion key files themselves for refine.bio supported
microarray platforms can be downloaded from [Zenodo](https://zenodo.org/record/1322711#.XLdJb-tKiXE).

```{r}
# Unzip hugene11st ID conversion file obtained from identifier refinery
id.conversion.file <- file.path(data_dir, "hugene11st.tsv")
if (!file.exists(id.conversion.file)) {
  unzip(paste0(id.conversion.file, ".zip"), exdir = data_dir)
}

# Read in the probe conversion info - this step may take some time
probe.conversion <- readr::read_tsv(id.conversion.file,
  progress = FALSE
) %>%
  dplyr::select("PROBEID", "ENSEMBL")

# Add annotation to imported author limma results
author.full.results <- author.full.results %>%
  # You may need to replace "affy_probe_ids" with the name of the column that
  # contains the probe IDs in your data
  dplyr::inner_join(probe.conversion, by = c("affy_probe_ids" = "PROBEID")) %>%
  dplyr::mutate("ENSEMBL" = as.factor(ENSEMBL)) %>%
  # Determine direction of fold change.
  # Note up-directed genes will be TRUE, down-genes will be FALSE
  # You may need to replace "logFC" with the name of the column that contains the
  # log fold change values in your data
  dplyr::mutate("direction" = logFC > 0) %>%
  # This is to make sure we don't get repeat probes
  dplyr::distinct()

# Take a look at the preview of this results table
author.full.results
```

### Summarize probe-level data into gene-level data

Because our refine.bio dataset has Ensembl gene IDs, and gene-level information
is generally more biologically meaningful and robust than probe-level information,
we will convert our Affymetrix probe IDs to Ensembl gene IDs.
We will find, however, that most Ensembl gene IDs have many probes that align to
them.
In order summarize probe-level into gene-level data, we will figure out for each
gene what the directions of the significant probes are.
Here we are only keeping adjusted p-values that are `< 0.05`.
This is a typical cutoff, but depending on your results and how long of gene
lists you would like to look at, you may need to adjust this.
Next, we will group the probes by their associated Ensembl gene IDs and count
how many probe sets are `up` and how many are `down` based on the `logical`
variable `direction` we made in the previous section.

```{r}
# Summarize the direction of significant probes for each gene
direction.summary <- author.full.results %>%
  # Keep only significant at 0.05 probes -- you may need to replace "adj.P.Val"
  # with the corresponding column name in your data that contains the adjusted P values.
  dplyr::filter(adj.P.Val < 0.05) %>%
  # Group probes by their associated
  dplyr::group_by(ENSEMBL) %>%
  dplyr::summarize(
    "up" = sum(direction),
    "down" = sum(!direction)
  )

# Print out this dataset
direction.summary
```

Now we have a count of how many probes are significant in each direction.
We will summarize these gene level probe summaries into two gene lists.
Some genes may have probe sets that are both up and down.
For this analysis, we will only keep genes in the significance lists if all the
probe sets are in the same direction.

```{r}
# Create an up-regulated genes list
author.up.genes <- direction.summary %>%
  # Upregulated genes are only those that have no significant downregulated
  # probe sets
  dplyr::filter(down == 0) %>%
  dplyr::pull(ENSEMBL) %>%
  as.character()

author.down.genes <- direction.summary %>%
  # Downregulated genes are only those that have no significant upregulated
  # probe sets
  dplyr::filter(up == 0) %>%
  dplyr::pull(ENSEMBL) %>%
  as.character()
```

## 4) Import and set up refine.bio processed data.

Data downloaded from refine.bio are in tab separated values (TSV) files and
include a metadata in a separate TSV file.
This data set was obtained from [refine.bio](https://www.refine.bio/).

```{r}
# Put name of data file
refine.bio.data.filename <- file.path(
  data_dir,
  "GSE37418.tsv" # Replace with your desired refine.bio processed data file
)

# Read in data TSV file
refine.bio.df <- readr::read_tsv(refine.bio.data.filename, progress = FALSE) %>%
  # Make the gene column the rownames so the gene names are out of the way for calculations
  tibble::column_to_rownames("Gene")
```

Set up the metadata for this set.

```{r}
# Metadata file name
metadata.filename <- file.path(
  data_dir,
  "metadata_GSE37418.tsv" # Replace with name of metadata file relevant to your refine.bio processed data file
)

# Read in metadata TSV file
metadata <- readr::read_tsv(metadata.filename) %>%
  # Get rid of columns that have only NAs
  dplyr::select(-which(apply(is.na(.), 2, all)))

# Print out metadata so we can get an idea of what we have
metadata
```

Let's reformat our variable `subgroup` that we will be using for testing.
To make the comparisons more reasonably similar, we will get rid of samples
that *aren't* SHH, Group3, or Group4.

```{r}
# Reformat subgroup names to make Group 3 and 4 into "NonSHH"
metadata$subgroup <- gsub("G3|G4", "NonSHH", metadata$subgroup)

# Check out subgroup summary
summary(as.factor(metadata$subgroup))
```

To make this dataset more comparable to our imported dataset (which only has SHH,
Group3, or Group4), we will get rid of samples that *aren't* in one of these
classifications.

```{r}
# Only keep samples that are SHH or Non-SHH (which is now both Group3 and Group4)
metadata <- metadata %>%
  dplyr::filter(subgroup %in% c("SHH", "NonSHH"))
```

Make refine.bio metadata and data are same samples, in the same order.

```{r}
# Reorder/filter data to match metadata
refine.bio.df <- refine.bio.df[, match(
  metadata$refinebio_accession_code,
  colnames(refine.bio.df)
)]

# Check if they are the same:
all.equal(colnames(refine.bio.df), metadata$refinebio_accession_code)
```

## 5) Perform differential expression analysis on refine.bio data

Here we are using the same model and set up as we used for the Northcott data
in [`author_processed_DE`](./author_processed_DE.nb.html).
`limma` needs a numeric design matrix to signify which samples belong to which
groups you are testing.
Note that we are using the medulloblastoma subgroups as our variable for testing.
This is the same type of variable we used for our imported author processed
dataset (See [`author_processed_DE`](./author_processed_DE.nb.html)).

```{r Set up design matrix}
# Make subgroup a factor
metadata$subgroup <- as.factor(metadata$subgroup)

# Create the design matrix
des.mat <- model.matrix(~ metadata$subgroup)

# Make the column name more informative
colnames(des.mat)[2] <- paste0(levels(metadata$subgroup), collapse = "vs")
```

## 6) Apply linear model

Here we will fit our data to a model using our design matrix and then apply
empirical Bayes for smoothing.
This same kind of model was applied to the Northcott data in [`author_processed_DE`](./author_processed_DE.nb.html)).

```{r Apply linear model}
# Initial fit
fit <- lmFit(refine.bio.df, design = des.mat)

# eBayes for smoothing
fit <- eBayes(fit)
```

Obtain summary statistics and apply Benjamini-Hochberg multiple testing
correction to the p-values.

```{r}
# Apply multiple testing correction and obtain stats
refine.bio.results <- topTable(fit, number = nrow(refine.bio.df)) %>%
  tibble::rownames_to_column("ENSEMBL")

# Print out stats
refine.bio.results
```

Write these results to an output file.

```{r  Write results}
readr::write_tsv(refine.bio.results, file.path(
  results_dir,
  "GSE37418_limma_results.tsv" # Replace with relevant output file name
))
```

### Create up/down-regulated gene lists for refine.bio results

Here we are using an adjusted p-value cutoff of `0.05`.
We are not applying a magnitude cutoff for `logFC` but using logFC just to
summarize direction of regulation.
Depending on your own data, you may want to adjust these.
For instance, you may only be interested in genes with high fold change.

```{r}
# Create up-regulated gene list for refinebio
refine.bio.up.genes <- refine.bio.results %>%
  # Apply cutoffs for p values and fold change direction to be up
  dplyr::filter(adj.P.Val < 0.05, logFC > 0) %>%
  # Reduce this to a gene ID vector list
  dplyr::pull(ENSEMBL) %>%
  as.character()

# Create down-regulated gene list for refinebio
refine.bio.down.genes <- refine.bio.results %>%
  dplyr::filter(adj.P.Val < 0.05, logFC < 0) %>%
  dplyr::pull(ENSEMBL) %>%
  as.character()
```

## 7) Double check results directions with a boxplot

Let's pick a couple genes from the Robinson and Northcott results to double check
our output.
We previously did this same test for the Northcott data in the in [`author_processed_DE`](./author_processed_DE.nb.html).

```{r}
# Extract a gene's data from the original refine.bio data.frame
test.gene <- refine.bio.df %>%
  tibble::rownames_to_column("genes") %>%
  dplyr::filter(genes == refine.bio.up.genes[12]) %>%
  dplyr::select(-genes) %>%
  t()

# Make a dataframe with the group information
box.plot <- data.frame(test.gene, subgroup = metadata$subgroup) %>%

  # Use ggplot2 to make a boxplot from this info
  ggplot(., aes(x = subgroup, y = test.gene)) +
  geom_boxplot() +
  theme_classic() +
  ggtitle(paste("refine.bio up gene test:", refine.bio.up.genes[12]))

# Print the plot here
box.plot
```

## 8) Set up exploration of overlap of the datasets differential expression

Here we will set up the data for making two Venn diagrams: one for overlap
between the datasets for up-regulated genes, the other for down-regulated genes.
Determine overlap in up-regulated genes between datasets.

```{r}
# Make into named list
up.genes <- list(
  "refine.bio" = refine.bio.up.genes,
  "author" = author.up.genes
)

# Make into named list
down.genes <- list(
  "refine.bio" = refine.bio.down.genes,
  "author" = author.down.genes
)
```

Determine overlap in gene sets between datasets.

```{r}
# Calculate overlap of these gene lists:
up.overlap <- VennDiagram::calculate.overlap(up.genes)

# Calculate overlap of these gene lists:
down.overlap <- VennDiagram::calculate.overlap(down.genes)
```

## 9) Make Venn diagrams

### Overlap between up-regulated genes

Make Venn diagram of up-regulation results and save to PNG.

```{r}
# Make filename to save plot as
up.plot.file <- file.path(
  plots_dir,
  "up-regulated_venn_northcott_vs_robinson.png"
)

# Make the Venn diagram
grid::grid.newpage()
up.venn.plot <- VennDiagram::draw.pairwise.venn(
  area1 = length(up.genes[[1]]),
  area2 = length(up.genes[[2]]),
  cross.area = length(up.overlap[[3]]),
  category = c("refine.bio", "author"),
  fill = c("blue", "yellow"),
  cex = 2,
  cat.cex = 1.5,
  cat.dist = c(-0.12, -0.13),
  cat.pos = c(170, 65),
  ext.pos = 0,
  ext.dist = -0.01,
  ext.length = .8,
  ext.line.lwd = 2,
  ext.line.lty = "dashed"
)
grid::grid.draw(up.venn.plot) # Draw plot
```

Save up-regulated genes Venn to a PNG.

```{r}
png(up.plot.file)
grid::grid.draw(up.venn.plot)
dev.off()
```

### Overlap between down-regulated genes

Repeat the process for down-regulation results.

```{r}
# Make filename to save plot as
down.plot.file <- file.path(
  plots_dir,
  "down-regulated_venn_northcott_vs_robinson.png"
)

# Make the Venn diagram
grid::grid.newpage()
down.venn.plot <- VennDiagram::draw.pairwise.venn(
  area1 = length(down.genes[[1]]),
  area2 = length(down.genes[[2]]),
  cross.area = length(down.overlap[[3]]),
  title = "Down-regulated Genes",
  category = c("refine.bio", "author"),
  fill = c("blue", "yellow"),
  cex = 2,
  cat.cex = 1.5,
  cat.dist = c(-0.11, -0.035),
  cat.pos = c(240, 120),
  ext.pos = 0,
  ext.dist = -0.01,
  ext.length = .8,
  ext.line.lwd = 2,
  ext.line.lty = "dashed"
)
grid::grid.draw(down.venn.plot) # Draw plot
```

Save down-regulated genes Venn to a PNG.

```{r}
png(down.plot.file)
grid::grid.draw(down.venn.plot)
dev.off()
```

## 10) Write overlapped gene list to TSV
Using `calculate.overlap`, we obtained a list of genes that were changed in both
datasets.
If we want to extract the list of genes that are in both dataset, that is the
third vector in the output object so we use `[[3]]`.

```{r}
# Write the up-regulated gene overlap lists to TSV file
data.frame(up.overlap[[3]]) %>%
  readr::write_tsv(file.path(results_dir, "up-regulated_overlap_gene_list.tsv"))

# Do the same but for the down-regulated gene overlap list
data.frame(down.overlap[[3]]) %>%
  readr::write_tsv(file.path(results_dir, "down-regulated_overlap_gene_list.tsv"))
```

Print session info:

```{r}
sessioninfo::session_info()
```
