---
title: "Clustering Data - Microarray"
author: "ALSF CCDL - Candace Savonen"
output:   
  html_notebook: 
    toc: true
    toc_float: true
    number_sections: true
bibliography: ../../references.bib  
---

*Purpose*: This notebook illustrates one way that you can use harmonized data 
from refine.bio in downstream analyses.

## Install libraries
This script uses the bioconductor R package [ComplexHeatmap](https://jokergoo.github.io/ComplexHeatmap-reference/book/) for clustering and 
creating a heatmap. [@ComplexHeatmap]

```{r Install and attach the ComplexHeatmap library}
if (!("ComplexHeatmap" %in% installed.packages())) {
  # Install ComplexHeatmap
  BiocManager::install("ComplexHeatmap", update = FALSE)
}
```

Attach the `ComplexHeatmap` library:

```{r}
# Attach the library
library(ComplexHeatmap)

# Magrittr pipe
`%>%` <- dplyr::`%>%`

# Set the seed so our results are reproducible:
set.seed(12345)
```

Create output folder.

```{r}
# Define the file path to the plots directory
plots_dir <- "plots" # Replace with path to desired output plots directory

# Create the plots folder if it doesn't exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

## Import and set up data
Data downloaded from refine.bio include a metadata tab separated values ("tsv")
file and a data tsv file. This chunk of code will read the both tsv files and 
add them as data.frames to your environment.

```{r Import data from .tsv files}
# Read in metadata tsv file
metadata <- readr::read_tsv(file.path("data", # Replace with path to your metadata file
                                      "metadata_GSE12955.tsv" # Replace with the name of your metadata file
                                      ))

# Read in data tsv file
df <- readr::read_tsv(file.path("data", # Replace with path to your data file
                                "GSE12955.tsv" # Replace with the name of your data file
                                )) %>%
  tibble::column_to_rownames('Gene')
```

Let's ensure that the metadata and data are in the same sample order. 

```{r}
# Make the data in the order of the metadata
df <- df %>% dplyr::select(metadata$geo_accession)

# Check if this is in the same order
all.equal(colnames(df), metadata$geo_accession)
```

## Choose genes of interest
Although you may want to create a heatmap including all of the genes in the set,
alternatively, the heatmap could be created using only genes of interest. 
For this example, we will sort genes by variance, but there are many alternative
criterion by which you may want to sort your genes <i>eg</i> fold change,
t-statistic, membership to a particular gene ontology, so on. 

```{r Choose genes}
# Calculate the variance for each gene
variances <- apply(df, 1, var)

# Determine summary statistics for gene variances
sum.stats.var <- summary(variances)

# Subset the data choosing only genes whose variances are in the upper quartile
df.by.var <- df[which(variances > sum.stats.var[5]), ]
```

## Create a heatmap
To further customize the heatmap, see the vignettes on Bioconductor for a guide 
at this [link](https://bioconductor.org/packages/devel/bioc/vignettes/ComplexHeatmap/inst/doc/s1.introduction.html).

```{r Create a heatmap}
# Reference the ComplexHeatmap guide for further customizing your heatmap
browseVignettes("ComplexHeatmap")

# Create an annotation that labels samples' groups by color
annot <-
  HeatmapAnnotation(df = data.frame(Groups = rep(c("SP", "MP"), # Replace with group names relevant to your data
                                                 each = 2)),
                    col = list(Groups = c("SP" = "green", "MP" = "orange"))) # Be sure to change the group names here as well -- a list of possible colors that can be used to replace the colors used here can be found at the following link: http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf

# Create the heatmap object
heatmap <- Heatmap(df.by.var, 
        name = "Gene_Expression",
        show_row_names = FALSE,
        show_row_dend = FALSE,   # Can show the gene/row cluster if this is 
        #changed to TRUE
        column_dend_height = unit(4, "cm"),
        bottom_annotation = annot) # assign the previously made annotation object

# Print out the heatmap
heatmap
```

## Save heatmap as a png
You can easily switch this to save to a jpeg or tiff by changing the function 
and file name within the function to the respective file suffix.

```{r Save heatmap as a png}
# Open a png file
png(file.path(plots_dir, 
              "HeatmapGSE12955.png" # Replace file name with a relevant output plot name
              ))

# Print your heatmap
heatmap

# Close the png file:
dev.off()
```

Print session info:

```{r Print session info}
# Print session info 
sessionInfo()
```
