---
title: "PCA"
author: "ALSF CCDL - Candace Savonen"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

*Purpose*: This notebook illustrates how to perform Principal Component 
Analysis and plot the scores using ggplot2. 

Recommended articles on PCA:   
- [Overall Explanation by Matt Brems](https://towardsdatascience.com/a-one-stop-shop-for-principal-component-analysis-5582fb7e0a9c)  
- [A visual explanation of PCA](http://setosa.io/ev/principal-component-analysis/)  
  
## 1) Set up libraries

```{r}
# Magrittr pipe
`%>%` <- dplyr::`%>%`

# Load ggplot2 for graphing
library(ggplot2)

# Make a results directory if it isn't created yet
if (!dir.exists("results")) {
  dir.create("results")
}
```

## 2) Import and set up data
Data downloaded from refine.bio are in tab separated values ("tsv") files and 
include a metadata in a separate tsv file. 
These data set was obtained from [refine.bio](https://www.refine.bio/). 

```{r}
# Put name of data file
data.filename <- file.path("data", "GSE71270.tsv")

# Read in data tsv file
df <- readr::read_tsv(data.filename, progress = FALSE) %>% 
  # Make the gene column the rownames so the gene names are out of the way for calculations
  tibble::column_to_rownames('X1') 
```

Import the metadata for this dataset

```{r}
# Metadata file name
metadata.filename <- file.path("data", "metadata_GSE71270.tsv")

# Read in metadata tsv file
metadata <- readr::read_tsv(metadata.filename) %>% 
  # Get rid of columns that have only NAs
  dplyr::select(-which(apply(is.na(.), 2, all)))

# Print out metadata so we can get an idea of what we have
metadata
```

## 3) Perform Principal Components Analysis.

```{r}
# Calculate PCA
pca <- prcomp(t(df))

# Make into dataframe for plotting PC1 and PC2 with ggplot2
pca.df <- data.frame(pca$x[, 1:2]) %>% 
  # Turn samples_ids stored as rownames into column
  tibble::rownames_to_column("refinebio_accession_code") %>%
  # Bring the metadata into this df; match by sample ids
  dplyr::inner_join(metadata, by = 'refinebio_accession_code')
```

Plot these PC scores for the first two principal components.
Label the data points based on their genotype group.

```{r}
# Make a scatterplot with ggplot2
ggplot(pca.df, aes(x = PC1, y = PC2, color = `genotype/variation`)) +
  geom_point() + 
  theme_classic()
```

## 4) Take a look at the gene loadings. 

```{r}
# Taking a look at PC1's biggest loadings:
head(sort(pca$rotation[, 1], decreasing = TRUE))
```

```{r}
# Taking a look at PC2's biggest loadings:
head(sort(pca$rotation[, 2], decreasing = TRUE))
```

## 5) Save PC scores to tsv file.

```{r}
pca.df %>% as.data.frame() %>% 
  tibble::rownames_to_column("sample") %>%
  readr::write_tsv(file.path("results", "PCA_scores.tsv"))
```

Session info: 

```{r}
# Print session info 
sessionInfo()
```
