---
title: "PCA Visualization"
author: "ALSF CCDL - Candace Savonen"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

*Purpose*: This notebook illustrates how to perform Principal Component 
Analysis and plot the scores using ggplot2. 

Recommended articles on PCA:   
- [Overall Explanation by Matt Brems](https://towardsdatascience.com/a-one-stop-shop-for-principal-component-analysis-5582fb7e0a9c)  
- [A visual explanation of PCA](http://setosa.io/ev/principal-component-analysis/)  
  
## 1) Set up

```{r}
# Set seed
set.seed(12345)

# Magrittr pipe
`%>%` <- dplyr::`%>%`

# Load ggplot2 for graphing
library(ggplot2)
```

Set up output directories

```{r}
# Define the file path to the results directory
results_dir <- "results"  # Replace with path to desired output results directory

# Make a results directory if it isn't created yet
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

# Define the file path to the plots directory
plots_dir <- "plots"  # Replace with path to desired output plots directory

# Make a plots directory if it isn't created yet
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

## 2) Import and set up data
Data downloaded from refine.bio are in tab separated values ("tsv") files and 
include a metadata in a separate tsv file. 
These data set was obtained from [refine.bio](https://www.refine.bio/). 

```{r}
# Put name of data file
data.filename <- file.path("data", # Replace with path to your data file
                           "GSE37382.tsv") # Replace with the name of your data file

# Unzip the file and direct it to the data folder
unzip(paste0(data.filename, ".zip"),
      exdir = "data" # Replace with path to your data file
      )

# Read in data tsv file
df <- readr::read_tsv(data.filename, progress = FALSE) %>% 
  # Make the gene column the rownames so the gene names are out of the way for calculations
  tibble::column_to_rownames('Gene') 
```

Import the metadata for this dataset

```{r}
# Metadata file name
metadata.filename <- file.path("data", # Replace with path to your metadata file
                               "metadata_GSE37382.tsv" # Replace with the name of your metadata file
                               ) 

# Read in metadata tsv file
metadata <- readr::read_tsv(metadata.filename) %>% 
  # Get rid of columns that have only NAs
  dplyr::select(-which(apply(is.na(.), 2, all)))

# Print out metadata so we can get an idea of what we have
metadata
```

## 3) Perform Principal Components Analysis.

```{r}
# Calculate PCA
pca <- prcomp(t(df))

# Make into dataframe for plotting PC1 and PC2 with ggplot2
pca.df <- data.frame(pca$x[, 1:2]) %>% 
  # Turn samples_ids stored as rownames into column
  tibble::rownames_to_column("refinebio_accession_code") %>%
  # Bring the metadata into this df; match by sample ids
  dplyr::inner_join(dplyr::select(metadata, refinebio_accession_code, 
                                  histology, subgroup), 
                    by = 'refinebio_accession_code')
```

Plot these PC scores for the first two principal components.
Label the data points based on their genotype group.

```{r}
# Make a scatterplot with ggplot2
pca.plot <- ggplot(pca.df, aes(x = PC1, y = PC2, color = subgroup)) +
  geom_point() + 
  theme_classic()

# Print out plot here
pca.plot
```

## 4) Save plot to png.

```{r}
# Save as a png
ggsave(file.path(plots_dir,
                 "pca_scatterplot.png" # Replace with a relevant output plot name
                 ))
```

## 5) Save PC scores to tsv file.

```{r}
pca.df %>% as.data.frame() %>% 
  readr::write_tsv(file.path(results_dir,
                             "pca_scores.tsv" # Replace with a relevant output name
                             ))
```

Session info: 

```{r}
# Print session info 
sessionInfo()
```
