---
title: "UMAP Visualization"
author: "ALSF CCDL - Candace Savonen"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

*Purpose*: This notebook illustrates how to perform Uniform Manifold Approximation
and plot the coordinates using ggplot2. 

Recommended articles on UMAP:   
- [How UMAP works with visuals](https://umap-learn.readthedocs.io/en/latest/how_umap_works.html)  
- [Original Paper on UMAP](https://arxiv.org/abs/1802.03426) by McGinnes, Healy & Melville.  

## 1) Set up libraries
Install [`umap` R package](https://cran.r-project.org/web/packages/umap/vignettes/umap.html) 
if you have not installed it previously. 

```{r eval = FALSE}
install.packages("umap")
```

```{r}
# Magrittr pipe
`%>%` <- dplyr::`%>%`

# Load ggplot2 for graphing
library(ggplot2)

# Make a results directory if it isn't created yet
if (!dir.exists("results")) {
  dir.create("results")
}
```

## 2) Import and set up data
Data downloaded from refine.bio are in tab separated values ("tsv") files and 
include a metadata in a separate tsv file. 
These data set was obtained from [refine.bio](https://www.refine.bio/). 

```{r}
# Put name of data file
data.filename <- file.path("data", "GSE37382.tsv")

# Unzip the file and direct it to the data folder
unzip(paste0(data.filename, ".zip"), exdir = "data")

# Read in data tsv file
df <- readr::read_tsv(data.filename, progress = FALSE) %>% 
  # Make the gene column the rownames so the gene names are out of the way for calculations
  tibble::column_to_rownames('X1') 
```

Import the metadata for this dataset

```{r}
# Metadata file name
metadata.filename <- file.path("data", "metadata_GSE37382.tsv")

# Read in metadata tsv file
metadata <- readr::read_tsv(metadata.filename) %>% 
  # Get rid of columns that have only NAs
  dplyr::select(-which(apply(is.na(.), 2, all)))

# Print out metadata so we can get an idea of what we have
metadata
```

## 3) Perform Uniform Manifold Approximation.

```{r}
# Calculate umap
umap.dat <- umap::umap(t(df))

# Make into dataframe 
umap.df <- data.frame(umap.dat$layout) %>% 
  # Turn samples_ids stored as rownames into column
  tibble::rownames_to_column("refinebio_accession_code") %>%
  # Bring the metadata into this df; match by sample ids
  dplyr::inner_join(dplyr::select(metadata, refinebio_accession_code, 
                                  histology, subgroup), 
                    by = 'refinebio_accession_code')
```

Plot the UMAP coordinates.
Label the data points based on their subgroup.

```{r}
# Make a scatterplot with ggplot2
ggplot(umap.df, aes(x = X1, y = X2, color = subgroup)) +
  geom_point() + 
  theme_classic()
```

## 5) Save UMAP coordinates to tsv file.

```{r}
umap.df %>% as.data.frame() %>% 
  readr::write_tsv(file.path("results", "UMAP_scores.tsv"))
```

Session info: 

```{r}
# Print session info 
sessionInfo()
```
