---
title: "Obtaining Annotation for Ensembl IDs"
author: "ALSF CCDL - Candace Savonen"
date: "9/17/2018"
output: html_document
---

*Purpose*: This notebook is an example annotation workflow for data obtained from refine.bio
using Bioconductor annotation packages.

Ensembl IDs can be used to obtain various different annotations at the 
gene/transcript level.
Although this example script uses Ensembl IDs from Zebrafish, (<i>Danio rerio</i>),
to obtain gene symbols this script can be easily converted for use with different 
species or annotation types <i>eg</i> protein IDs, gene ontology, accession 
numbers. 

<b>For different species</b>, wherever the abbreviation `org.Dr.eg.db` or `Dr` is 
written, it must be replaced with the respective species abbreviation <i>eg</i>
for<i> Homo sapiens</i> `org.Hs.eg.db` or `Hs` would be used.
A full list of the annotation R packages from Bioconductor is at this 
<a href="http://bioconductor.org/packages/release/data/annotation"> link</a>.

<b>For different types of annotation</b>, other annotation objects besides
`org.Dr.egSYMBOL` can be used for mapping.
For a full list annotation objects from a package this command can be used: 
`ls("package::org.Dr.eg.db")`. 
In this example, wherever `SYMBOL` is written, it would need to be replaced 
with the name of the desired annotation such as `ACCNUM` for accession numbers.

## 1) Install libraries with species-specific annotation package
```{r Install libraries}
# Install the Zebrafish package
source("https://bioconductor.org/biocLite.R")
biocLite("org.Dr.eg.db", suppressUpdates = TRUE)

# Attach the annotation library 
library(org.Dr.eg.db)
```

## 2) Import refine.bio data that has Ensembl identifiers
```{r Import data}
# Ensembl IDs are in the first column and will be read in as the rownames for 
# this data frame
df <- read.delim("../data/SRP056136.tsv", row.names = 1, 
                 stringsAsFactors = FALSE)
```

## 3) Map Ensembl IDs to gene symbols
*Note*: This code chunk will return a message that indicates that there are many
mappings between our keys and columns but this is okay and is what we expect 
since genes and transcripts do not have a 1:1 relationship
```{r Match Ensembl IDs}
# Map ensembl IDs to their associated gene symbols
symbol.df <- data.frame('gene.symbols' = mapIds(org.Dr.eg.db, keys = rownames(df), 
                                  column = "SYMBOL", keytype = "ENSEMBL"))

# Merge gene symbol annotation with data frame
# Keep Ensembl IDs as row names using transform
annot.df <- transform(merge(symbol.df, df, by = "row.names"),
                        row.names = Row.names, Row.names = NULL)
```

## 4) Write to file
```{r Write to file}
# Write annotated data frame to output file
write.table(annot.df, file = "SRP056136wGeneSymbols.tsv", sep = "\t",
            quote = FALSE)
```

```{r Print session info}
# Print session info 
sessionInfo()
```
